Live Functional Programming with Typed Holes

### 2018-04-17 examples.tex
near-final intro
sec2!

% !TEX root = hazelnut-dynamics.tex
%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
\newcommand{\examplesSec}{\HazelnutLive, By Example}
\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:examples}

@@

We introduce \HazelnutLive{} in this section using several examples,
demonstrating how running incomplete programs---with missing expressions and/or
type-inconsistent expressions---can be useful.
%
We will also demonstrate how running even complete programs with subexpressions
marked as holes may be useful in debugging.

\HazelnutLive{} programs are written in OCaml \cite{leroy03:_ocaml}, using the
alternative Reason syntax \cite{reason-what} extended to support holes.
%
As shown in \autoref{fig:grades-example}, a \HazelnutLive{} notebook is a
literate program \cite{knuth1984literate}: rich text comments appear at the top
level and code appears within cells embedded into these comments.
%
The reader is encouraged to work through these examples in our prototype
implementation or watch video demonstrations, both available in
\suppMaterials{}.\footnote{
%
The programs in this section are written using minor syntactic conveniences that
are not currently supported in our implementation.
%
For example, the implementation of recursive types in our current implementation
requires explicit \texttt{fold} and \texttt{unfold} expressions; these can be
inferred in standard ways (by pairing them with constructors and deconstructors
for algebraic datatypes) in future work.
%
More details about other minor syntactic differences are described in
\suppMaterials{}.
%
}

@@

\paragraph{Recap}
%
In the above four subsections, respectively, we considered programs with:
%
(1) empty expression holes,
%
(2) non-empty expression holes for ill-typed expressions,
%
(3) type holes, and
%
(4) non-empty expression holes for well-typed expressions.
%
Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
that allows programs with combinations of these kinds of incompleteness to be
evaluated.

>>>>

% !TEX root = hazelnut-dynamics.tex
%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
\newcommand{\examplesSec}{Live Functional Programming with Typed Holes in Hazel}
\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:examples}

\input{fig-grades-example}

@@

We begin with an example-driven overview of our proposed approach  within the \Hazel programming environment \citet{HazelnutSNAPL}. \Hazel is an open source web lab notebook environment based roughly on IPython/Jupyter \cite{Perez:2007:ISI:1251563.1251831}. The language of \Hazel is tracking toward parity with \Elm, a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}. We assume familiarity with the basic concepts of ML/Elm (see \url{http://elm-lang.org/}).

Both the user interface and the language of \Hazel are under active development. For the sake of 
exposition, we have ``post-processed'' the examples in this section to make use of several convenient language mechanisms from \Elm~that are not yet implemented in \Hazel as of this writing. These mechanisms are orthogonal to the contributions of this paper. We have implemented all of the live programming features described in this section, both in a branch of \Hazel and in the implementation of the core calculus that we describe in Sec.~\ref{sec:implementation} and include in the supplement.

@@

% \paragraph{Recap}
% %
% In the above four subsections, respectively, we considered programs with:
% %
% (1) empty expression holes,
% %
% (2) non-empty expression holes for ill-typed expressions,
% %
% (3) type holes, and
% %
% (4) non-empty expression holes for well-typed expressions.
% %
% Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
% that allows programs with combinations of these kinds of incompleteness to be
% evaluated.

### 2018-04-17 fig-grades-example.tex
sec2!

\begin{figure}[t]

\includegraphics[width=1.01\textwidth,interpolate=false]{images/grades-cell-mockup.png}

% %% TODO once the code above is removed, scale up the screenshots
% \includegraphics[scale=0.20]{images/hazel-placeholder-0.png}

@@

% \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}

\caption{Hazel mockups for Example 1a.}
\label{fig:grades-example}
\end{figure}

>>>>

\begin{figure}[t]
\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=0.85\textwidth,interpolate=false]{images/grades-cell-mockup.png}
\caption{Evaluating an incomplete functional program past the first hole}
\label{fig:grades-cell-mockup}
\end{subfigure}

\vspace{10px}

\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-1.png}
~${}^\blacktriangleright$
\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-2.png}
~${}^\blacktriangleright$
\includegraphics[width=0.3\textwidth,interpolate=false]{images/grades-sidebar-3.png}
\caption{The programming environment communicates relevant static \emph{and} dynamic information in a sidebar.}
\label{fig:grades-sidebar}
\end{subfigure}
% %% TODO once the code above is removed, scale up the screenshots
% \includegraphics[scale=0.20]{images/hazel-placeholder-0.png}

@@

% \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
\vspace{5px}

\caption{Example 1: Grades}
\label{fig:grades-example}
\vspace{-8px}
\end{figure}

### 2018-04-17 hazelnut-dynamics.tex
pass over Section 4; copy edits and reorganization
sec2!

%% Cyrus packages
\usepackage{microtype}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{leftidx}
\usepackage{todonotes}
\usepackage{xspace}

\usepackage{listings}%
\lstloadlanguages{ML}
\lstset{tabsize=2, 
basicstyle=\footnotesize\ttfamily, 
% keywordstyle=\sffamily,
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{purple},
mathescape=false,escapechar=\#,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
morekeywords={string, float, int},
classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount
}
\let\li\lstinline
\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}

>>>>

%% Cyrus packages
\usepackage{microtype}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{leftidx}
\usepackage{todonotes}
\usepackage{xspace}
\usepackage{wrapfig}
\usepackage[export]{adjustbox}% http://ctan.org/pkg/adjustbox

\usepackage{listings}%
\lstloadlanguages{ML}
\lstset{tabsize=2, 
basicstyle=\footnotesize\ttfamily, 
% keywordstyle=\sffamily,
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{purple},
mathescape=false,escapechar=\#,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
morekeywords={string, float, int},
classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
moredelim=**[is][\color{red}]{SSTR}{ESTR}
}
\newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}

### 2018-04-17 related-work.tex
sec2!

As already discussed in Track 1, type holes are closely related to the unknown
types of gradual type theory \cite{Siek06a}. This has implications for our
dynamic semantics as well---we need to be able to run programs where type holes
have obscured a type inconsistency, \eg{}~\li{let f(g : }\ \li{) => g 3 in f 4}.
This term is well-typed according to the static semantics of Track 1 because the
argument \li{g} of \li{f} has an unknown type, and so it can be applied to
\li{3}, but when \li{f} is applied to \li{4}, there is a problem because \li{4 3}
is meaningless. There are two ways to address the problem. The first is to
apply ML-style type inference to discover a more specific type schema for
\li{g}, \ie{}~that it must actually be an arrow type, and then mark the argument
\li{4} to \li{f} as a type inconsistency using a non-empty hole. We will
investigate this in Track 1. The second approach is to adopt run-time cast
insertions from research on gradual type theory. We will investigate this second
approach in Track 2.

@@

put Seidel here

\parahead{Debugging}

>>>>

As already discussed in Track 1, type holes are closely related to the unknown
types of gradual type theory \cite{Siek06a}. This has implications for our
dynamic semantics as well---we need to be able to run programs where type holes
have obscured a type inconsistency, \eg{}~\lstinline{let f(g : }\ \lstinline{) => g 3 in f 4}.
This term is well-typed according to the static semantics of Track 1 because the
argument \li{g} of \li{f} has an unknown type, and so it can be applied to
\li{3}, but when \li{f} is applied to \li{4}, there is a problem because \li{4 3}
is meaningless. There are two ways to address the problem. The first is to
apply ML-style type inference to discover a more specific type schema for
\li{g}, \ie{}~that it must actually be an arrow type, and then mark the argument
\li{4} to \li{f} as a type inconsistency using a non-empty hole. We will
investigate this in Track 1. The second approach is to adopt run-time cast
insertions from research on gradual type theory. We will investigate this second
approach in Track 2.

@@

Understanding and debugging static type errors is notoriously difficult,
particularly for novices.
%
A variety of approaches have been
proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
and explain type errors.
%
One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
that demonstrates a run-time failure, and then displays a compressed execution
trace to the user as a graph.
%
In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
non-empty holes) as far as possible, in particular, until they
would go wrong.
%
Thus, our operational semantics provides a similar benefit to the approach of
\citet{Seidel2016}.

\parahead{Debugging}
This approach is reminiscent of the workflow that debuggers make available using breakpoints \cite{fitzgerald2008debugging,DBLP:journals/jfp/TolmachA95}. 
The difference here is twofold. First, evaluation does not stop at each breakpoint and so it is straightforward to explore the \emph{space} of values that a variable takes. Second, breakpoints make available the values of a variable at a position in the evaluation trace. Hole closures, on the other hand, convey information from a syntactic position in the result of evaluation. The result is, by nature, a simpler object than the full evaluation trace that produced the result. These approaches are fundamentally complementary, rather than opposed.\todo{move to related work?}{}\todo{talk about tracing/recording}{}
